---
title: "Importing interlinearized corpora and dictionary into R"
author: "Sylvain Loiseau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing interlinearized corpora and dictionary into R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(interlineaR)
```


# Introduction

This package contains utility functions for reading in into R documents
in various file format containing interlinearized corpora or dictionary
produced by descriptive linguistics softwares.

Short selection of (very preliminary draft of a) dictionary and an
interlinearized corpus of the Tuwari (Walioic (papuan), PNG) language
[http://glottolog.org/resource/languoid/id/tuwa1242]
are provided in the exampleData directory.

# Reading EMELD XML interlinear corpus

XML vocabulary introduced in Baden Hughes, Steven Bird and Catherine Bow, _Encoding and Presenting Interlinear Text Using XML Technologies_, [http://www.aclweb.org/anthology/U03-1008], it is used by SIL Fieldworks as an export format.

```{r}
corpuspath <- system.file("exampleData", "tuwariInterlinear.xml", package="RFlex")
corpus <- read.emeld(corpuspath, vernacular.languages="tww")
```

# Reading LIFT XML dictionary

This XML vocabulary has been introduced by SIL : [http://code.google.com/p/lift-standard] and is used by SIL Fieldworks as an export format.

```{r}
dicpath <- system.file("exampleData", "tuwariDictionary.lift", package="RFlex")
dictionary <- read.lift(dicpath, language.code="tww")
```

